---
title: Overview
author: David Kane and Ansh Pathak
tutorial:
  id: 99-overview
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: 'Overview tutorial for entire book'
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(tidycensus)
library(tidyverse)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
options(tutorial.exercise.timelimit = 600, 
        tutorial.storage = "local") 
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

<!-- DK: Replace primer.data with tidycensus. Done -->

## Introduction
###


### Exercise 1

Go to the [Census website](https://api.census.gov/data/key_signup.html) to request a Census API key. The "Organization Name" you provide can be anything, including your current or former school. This key will allows you to download, for free, US Census data. The key will be emailed to you, but it may take a bit of time. You can still proceed with this tutorial without it.

From the Console, run `Sys.getenv("CENSUS_API_KEY")`. CP/CR. 

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

Note that, because we have not done anything with your key yet --- in fact, the Census may not have even emailed it back to you --- this command should return `""`.

### 

And that is OK! We have not created the `CENSUS_API_KEY` environment variable yet. We will create this variable at the end of this tutorial.


### Exercise 2

Load the **tidyverse** package.


```{r introduction-2, exercise = TRUE}

```

```{r introduction-2-hint-1, eval = FALSE}
library(...)
```

```{r introduction-2-test, include = FALSE}
library(tidyverse)
```

### 

Aggregate data from the decennial US Census, American Community Survey, and other Census surveys are made available to the public at different enumeration units. 

### 

Enumeration units are geographies at which Census data are tabulated. They include both legal entities such as states and counties, and statistical entities that are not official jurisdictions but used to standardize data tabulation. 

### Exercise 3

Load the **tidycensus** package.

```{r introduction-3, exercise = TRUE}

```

```{r introduction-3-hint-1, eval = FALSE}
library(...)
```

```{r introduction-3-test, include = FALSE}
library(tidycensus)
```

###


## Texas Income



The smallest unit at which data are made available from the decennial US Census is the block, and the smallest unit available in the ACS is the block group, which represents a collection of blocks. Other surveys are generally available at higher levels of aggregation.


### Exercise 1

Create a Github repo called `Tidycensus-plots`. Make sure to click the "Add a README file" check box.

Connect the repo to a project on your computer using `File -> New Folder from Git ...`. Make sure to select the "Open in a new window" box. 

You need two Positon windows: this one for running the tutorial and the one you just created for writing your code and interacting with the Console.

Select `File -> New File -> Quarto Document ...`. Provide a title -- `"TexasIncome"` -- and an author (you). Render the document and save it as `TexasIncome.qmd`.

Create a `.gitignore` file with `_files` on the first line and then a blank line. Save and push.

In the Console, run:

```         
show_file(".gitignore")
```

If that fails, it is probably because you have not yet loaded `library(tutorial.helpers)` in the Console.

CP/CR.

```{r texas-income-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

### 

Professionals keep their data science work in the cloud because laptops fail.

### Exercise 2

<!-- XX: Switch primer.data for whichever package you get your data from. -->

In your QMD, put `library(tidyverse)` and  `library(tidycensus)` in a new code chunk. Press Ctrl/Cmd + Shift + K to render the file

Notice that the file does not look good because the code is visible and there are annoying messages. To take care of this, add `#| message: false` to remove all the messages in this `setup` chunk. Also add the following to the YAML header to remove all code echos from the HTML:

```         
execute: 
  echo: false
```

In the Console, run:

```         
tutorial.helpers::show_file("TexasIncome.qmd", start = -5)
```

CP/CR.

```{r texas-income-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

### 

Render again. Everything looks nice, albeit empty, because we have added code to make the file look better and more professional.

### Exercise 3

<!-- XX: This reason for this somewhat convoluted approach is that we want to give students lots of practice working in both the QMD World and the Console World. -->

Place your cursor in the QMD file on the `library(tidyverse)` line. Use `Cmd/Ctrl + Enter` to execute that line.

Note that this causes `library(tidyverse)` to be copied down to the Console and then executed. 

CP/CR.

```{r texas-income-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

<!-- XX: Report the source of the data, along with some background information about it. -->

### Exercise 4

In this exercise, you will use an AI assistant (such as ChatGPT) to generate R code that:

- Loads the `tidycensus` and `tidyverse` libraries.
- Uses `get_acs()` to download median household income (`B19013_001`) data for **all counties in Texas** for the year **2020**.
- Saves the data to a variable named `income_tx`.
- Includes `geometry = TRUE` so the data includes spatial information.

**Instructions:**

1. Ask an AI assistant to generate the required R code to do the above.
2. Copy and paste the AI-generated code into your Quarto document and run it (Press Cmd/Ctrl + Enter and send the chunk to the console).
3. Copy the output from your Console and paste it here (CP/CR).

```{r texas-income-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```


###

Now, replace the code it gave you with this code: 

```r
library(tidycensus)
library(tidyverse)

income_tx <- get_acs(
  geography = "county",
  variables = "B19013_001",
  state = "TX",
  year = 2020,
  geometry = TRUE
)
```

`get_acs()` is part of the tidycensus package and allows downloading American Community Survey (ACS) data. The `geometry = TRUE` argument returns spatial polygons, useful for maps and spatial analysis.

### Exercise 5

Now you will use AI to generate code that creates a plot of median household income in Texas counties.

The goal is to create a choropleth map using `ggplot2` showing the median income by county using the `income_tx` dataset.

**Instructions:**

1. Ask an AI assistant to generate R code that:
   - Uses `ggplot2` to create a choropleth map of median household income.
   - Uses `geom_sf()` to plot the spatial data.
   - Colors counties by the estimate column.
   - Adds appropriate titles and themes.
2. Copy and paste the AI-generated code into your Quarto document and run it.
3. Copy the output from your Console and paste it here (CP/CR).

```{r texas-income-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Now, replace the code it gave you with this code: 

```r
library(ggplot2)

income_map <- ggplot(income_tx) +
  geom_sf(aes(fill = estimate), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", name = "Median Income") +
  labs(`
    title = "Median Household Income by County, Texas (2020)",
    caption = "Data source: ACS 5-year estimates"`
  ) +
  theme_minimal()

income_map
```

### Exercise 6


1. In the Console, run the following command to display the last chunk of your `.qmd` file: CP/CR

tutorial.helpers::show_file("TexasIncome.qmd", chunk = "last")

```{r texas-income-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 15)
```

###

The `show_file()` function from tutorial.helpers is a convenient way to check the contents of files without leaving R. It helps confirm that your edits were saved properly.



`geom_sf()` plots spatial data stored as simple features (`sf` objects). The `fill` aesthetic maps color to median income. The `viridis` color scale provides perceptually uniform colors for better interpretation.


## California Bachelors Degree


Did you know?

Visualizing Maps with `geometry = TRUE`
Passing `geometry = TRUE` to `get_acs()` returns spatial geometry as an `sf` object, which works well with `ggplot2` for choropleth maps.


### Exercise 1

Select `File -> New File -> Quarto Document ...`. Provide a title -- `"CaliforniaBachelors"` -- and an author (you). Render the document and save it as `CaliforniaBachelors.qmd`.

In this exercise, we’ll get the percentage of adults with a bachelor’s degree or higher in each California county.

- Ask an AI assistant (like ChatGPT) to generate R code that:

Loads `tidycensus` and `tidyverse`
Uses `get_acs()` to get educational attainment variables for all California counties in 2020
Saves the result to a variable called `edu_ca`
Includes `geometry = TRUE`
Paste the AI-generated code into your Quarto file (`CaliforniaBachelors.qmd`). Run the code using `Cmd/Ctrl + Enter` in the QMD editor or render the file using `Ctrl/Cmd + Shift + K.`
Copy and paste the result from the Console here. CP/CR.

```{r california-bachelors-degree-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```


###

Now, replace the code it gave you with this code: 

```r
library(tidycensus)
library(tidyverse)

edu_ca <- get_acs(
  geography = "county",
  variables = c("B15003_001", "B15003_022", "B15003_023", "B15003_024", "B15003_025"),
  state = "CA",
  year = 2020,
  geometry = TRUE,
  summary_var = "B15003_001"
)
```


The `get_acs()` function is powerful for pulling American Community Survey (ACS) data. 
For educational attainment, we use `B15003_022` through `B15003_025` to sum all individuals with a bachelor’s degree or more, then divide by the total population (variable `B15003_001`). 
Geometry must be `TRUE` if we want to map later.

### Exercise 2

We’ll now make a choropleth map of bachelor’s degree attainment across California counties.

Ask an AI assistant to write code that:

Uses `mutate()` to calculate the percentage of the population with at least a bachelor’s degree
Pipes that into a `ggplot()`
Uses `geom_sf(aes(fill = percent))` and a `scale_fill_viridis_c()` to make it look nice
Adds labels and a title
Paste the AI-generated code into your Quarto document. Run it.

CP/CR.

```{r california-bachelors-degree-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Now, replace the code it gave you with this code: 

```r
edu_ca <- edu_ca %>%
  group_by(GEOID) %>%
  summarize(
    percent = 100 * sum(estimate[variable != "B15003_001"]) / unique(summary_est)
  )

ggplot(edu_ca) +
  geom_sf(aes(fill = percent)) +
  scale_fill_viridis_c(option = "C") +
  labs(title = "Percent of Adults with Bachelor's Degree or Higher in CA Counties (2020)",
       fill = "% with Degree") +
  theme_minimal()
```


`ggplot2` can handle spatial data directly using `geom_sf()`. Use `mutate()` to calculate percentages, and pipe that into `ggplot()` for a map.


### Exercise 3


1. In the Console, run the following command to display the last chunk of your `.qmd` file: CP/CR

tutorial.helpers::show_file("CaliforniaBachelors.qmd", chunk = "last")

```{r california-bachelors-degree-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 15)
```

###

The variables in `get_acs()` like `"B19013_001"` are codes that represent specific census survey questions or data points.

 You can use the `load_variables()` function from tidycensus to look up what each code means and explore available variables for a given year and dataset. 
 
 This helps you find exactly the data you need without guessing variable names.


## California Median Age


### Exercise 1

Select `File -> New File -> Quarto Document ...`. Provide a title -- `"CaliforniaAge"` -- and an author (you). Render the document and save it as `CaliforniaAge.qmd`.

In this exercise, you will collect median age data for all counties in California for the year 2020 using the `tidycensus` package.

The `get_acs()` function from `tidycensus` lets you download American Community Survey data. You specify geography, variables, state, and year. The variable `B01002_001` represents median age.

**Task:**

1. Ask an AI assistant (such as ChatGPT) to generate R code that:
   - Loads the `tidycensus` and `tidyverse` libraries,
   - Uses `get_acs()` to get median age (`B01002_001`) for all counties in California for 2020,
   - Saves the result to a variable called `age_ca`.

2. Copy the AI-generated code into your Quarto document and render the file (e.g., press Ctrl/Cmd + Shift + K).

3. Copy and paste the output from your Console here (CP/CR).
```{r california-median-age-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```


###

Now, replace the code it gave you with this code: 

```r
library(tidycensus)
library(tidyverse)

age_ca <- get_acs(
  geography = "county",
  variables = "B01002_001",
  state = "CA",
  year = 2020
)
age_ca
```


Census Variable Codes:
Each variable has an identifier. For example, `"B19013_001"` stands for median household income. Use the tidycensus variable lookup to find codes.

### Exercise 2

Now that you have the median age data for California counties, the goal is to create a bar plot showing median age by county.

In this exercise:

Ask an AI assistant to generate R code that:
Uses the `age_ca` data,
Creates a bar plot with `ggplot2` where counties (`NAME`) are on the y-axis, median age estimates on the x-axis,
Sorts counties by median age descending (highest to lowest),
Uses a horizontal bar chart (`coord_flip()`),
Adds a title and axis labels.

Copy the AI-generated code into your Quarto document and render it.
Copy and paste the console output here (CP/CR).

```{r california-median-age-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Now, replace the code it gave you with this code: 

```r
age_ca %>%
  mutate(NAME = fct_reorder(NAME, estimate, .desc = TRUE)) %>%
  ggplot(aes(x = NAME, y = estimate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Median Age by County in California (2020)",
    x = "County",
    y = "Median Age"
  ) +
  theme_minimal()
```


`coord_flip()` from ggplot2:
Flips x and y axes, often used to make horizontal bar charts easier to read.


### Exercise 3


1. In the Console, run the following command to display the last chunk of your `.qmd` file: CP/CR

tutorial.helpers::show_file("CaliforniaAge.qmd", chunk = "last")

```{r california-median-age-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 15)
```

###

`geom_col()` from `ggplot2`:
Creates bar charts where bar heights correspond to values in the data.

`theme_minimal()` from `ggplot2`:
Applies a clean, minimal theme to the plot for better readability.



```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
