---
title: An introduction to tidycensus
author: David Kane
tutorial:
  id: an-introduction-to-tidycensus
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: 'Tutorial for Chapter 2: An introduction to tidycensus'
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(tidycensus)
library(tidyverse)

knitr::opts_chunk$set(echo = FALSE)
options(tutorial.exercise.timelimit = 60, 
        tutorial.storage = "local") 


# TB question: does this code go here? library(tidycensus) # census_api_key("YOUR KEY GOES HERE", install = TRUE)
  
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

<!-- DK: This is a total mess! -->

## Introduction
### 

This tutorial covers [Chapter 2: An introduction to tidycensus](https://walker-data.com/census-r/an-introduction-to-tidycensus.html) from [*Analyzing US Census Data: Methods, Maps, and Models in R*](https://walker-data.com/census-r/index.html) by Kyle Walker. You will learn about the [**tidycensus**](https://walker-data.com/tidycensus/index.html) package and some of its key functions, including [`get_decennial()`](https://walker-data.com/tidycensus/articles/basic-usage.html) and [`get_acs()`](https://walker-data.com/tidycensus/articles/basic-usage.html). 

Why tidycensus? The US Census Bureau population data is a great source for researchers, and this packages facilitates your exploration of the Census data in the R environment.

## Getting started with tidycensus
### 

<!-- DK: Intro -->

### Exercise 1

Let's try accessing the Decennial Census database. In order to do this, we'll use the `get_decennial()` function. This allows us to access decennial Census data from the 2000, 2010, and 2020 decennial US Censuses.


First, type in the function `get_decennial()`.

```{r getting-started-with-tidycensu-1, exercise = TRUE}

```

```{r getting-started-with-tidycensu-1-hint-1, eval = FALSE}
get_decennial()
```

### 

Just calling the function is not enough, we need to specify which data we are looking to extract. This is done via the use of three arguments: `geography`, `variables`, and `year`.

`geography` argument: determines the scope of the data 
    `us`: the entire country 
    `county`: counties
    `state`: states
`variables` argument: defines which variables you want to use
`year` argument: defines the year that our data comes from

For this exercise, let's try to find the total American population by state in 2010. To do so, we need to access the variable `P001001`.

### Exercise 2

Under the previously in-puted function, set `geography` to `"state"`, `variables` to `"P001001"`, and `year` to `2010`.

```{r getting-started-with-tidycensu-2, exercise = TRUE}

```

```{r getting-started-with-tidycensu-2-hint-1, eval = FALSE}
get_decennial(... = "...", 
              variables = "...",
              ... = 2010)
```

### 

This should return a tibble with 52 rows and 4 columns.

Now, what if we want to see the data from 2020?


### Exercise 3

Copy and paste your code from above. Change the `year` argument from `2010` to `2020`. Change the `variables` argument to `"P1_001N"`, because the `"P001001"` was renamed starting in 2020.

```{r getting-started-with-tidycensu-3, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r getting-started-with-tidycensu-3-hint-1, eval = FALSE}
get_decennial(... = "...", 
              variables = "...",
              year = ...)
```

### 

Now you get a different data set than that from Exercise 2. By changing the `year` argument, we can specify when we want our data to be from. 


We should introduce another argument: `sumfile`, which stands for "summary file". 

Each version of the decennial US Censuses contain many summary files (`sf`), each containing different information.

*"sf1"*: Summary File 1, core demographic characteristics for Census geographies.
*"sf2"*: Summary File 2, information on a range of population and housing unit characteristics.
*"sf3"* and *"sf4"*: Summary File 3 and 4, income and occupation (2000 census only).

Island territories (except Puerto Rico) have their corresponding summary files.
*"as"*: Summary File for American Samoa
*"mp"*: Summary File for the Northern Mariana Islands
*"gu"*: Summary File for Guam
*"vi"*: Summary File for the US Virgin Islands.

*"pl"*: Redistricting summary file (2020 and 2010 census), include total population and housing units; race and ethnicity; voting-age population; and group quarters population. 

Note that by default, `get_decennial()` uses the argument `sumfile = "sf1"`. 

If you wish to extract data from other sumfiles, you need to specify this argument under the function. Now let's practice. 


We want to retrieve information on the *American Indian & Alaska Native population* by *state* from the *2020* decennial Census.

### Exercise 4

Run the `get_decennial()` function below setting `geography` to `"state"`, `variables` to `"P1_005N"`, `year` to `2020`, and `sumfile` to `"pl"`.

```{r getting-started-with-tidycensu-4, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r getting-started-with-tidycensu-4-hint-1, eval = FALSE}
get_decennial(geography = "...",
        variables      = "...",
        year = ...,
        sumfile = "..."
)
```

```{r getting-started-with-tidycensu-4-test, include = FALSE}
get_decennial(geography = "state",
        variables      = "P1_005N",
        year = 2020,
        sumfile = "pl"
)
```

### 

This retrieves information on the American Indian & Alaska Native population by state from the 2020 decennial Census. 

To learn more about the `sumfile`, refer to Walker's [book](https://walker-data.com/census-r/an-introduction-to-tidycensus.html#getting-started-with-tidycensus).

### Exercise 5

Just as `get_decennial()` gets information from the Decennial Census, `get_acs()` gets information from the **A**merican **C**ommunity **S**urvey. 

The ACS includes a wide variety of variables detailing characteristics of the US population not found in the decennial Census. 

The basic set-up of `get_acs()` is similar to that of `get_decennial()` as they all use the three arguments: `geography`, `variables` and `year`.

### Exercise 6

Let's look at the number of U.S. residents born in Mexico by state.

Run the `get_acs()` function, setting `geography` to  `"state"`, `variables` to `"B05006_150"` and `year` to `2021`.

```{r getting-started-with-tidycensu-6, exercise = TRUE}

```

```{r getting-started-with-tidycensu-6-hint-1, eval = FALSE}
get_acs(geography = "...",
        variables      = "...",
        year = ...
)
```

```{r getting-started-with-tidycensu-6-test, include = FALSE}
get_acs(geography = "state",
        variables = "B05006_150",
        year = 2021
)
```

### Exercise 7

Notice that the tibble produced has an `estimate` and a `moe` column rather than a `value` column like the Decennial Census data. 

`estimate` column: ACS estimate.
`moe` column: margin of error around that estimate.

### Exercise 8

What if you didn't specify 'year'? 

`get_acs()` will defaults to the most *recent five-year* ACS sample (ie. 2019-2023)

### Exercise 9

Another important argument is `survey`. 

This argument defaults to 5 year periods if not specified. 

We can change it to a one-year period rather than a default 5-year period by adding `survey = "acs1"`.

### Exercise 10

Run `get_acs()`, setting `geography` to `"state"`, `variables` to `"B05006_150"`, `survey` to `"acs1"`, and `year` to `2021`.


```{r getting-started-with-tidycensu-10, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r getting-started-with-tidycensu-10-hint-1, eval = FALSE}
get_acs(... = "...", 
        variables = "...", 
        survey = "...",
        ... = 2019)
```

```{r getting-started-with-tidycensu-10-test, include = FALSE}
get_acs(geography = "state", 
        variables = "B05006_150", 
        survey = "acs1",
        year = 2021)
```

### Exercise 11

What do you think this code entails?

Well, it yields data on the number of residents born in Mexico by state in the year of *just 2021* (instead of a 5-year period as earlier). 

### Exercise 12

You should have seen some `NA` in the resulted table of the previous exercise. This is R’s notation representing missing data. 

In this case, `NA` in the estimate column generally means that the estimate is too small for a given geography to be deemed reliable by the Census Bureau.

### Exercise 13

Notice that the GEOID column contains the [FIPS code](https://transition.fcc.gov/oet/info/maps/census/fips/fips.txt#:~:text=FIPS%20codes%20are%20numbers%20which,to%20which%20the%20county%20belongs.) of the area the data covers. This code is a standardized way to identify different geographical areas. It gets larger as the area under consideration narrows. The FIPS for Massachusetts is 25. For Middlesex County within Massachusetts, the FIPS is 25017.

## Geography and variables
### 

<!-- TB: start from here -->

### Exercise 1

Rather than using the `variables` argument, you can supply a table name to the `table` parameter in `get_acs()`; this will return data for every variable in that table. 

Run `get_acs()` below setting `geography` to `"state"`, `table` to `"B01001"`, and `year` to `2021`

```{r geography-and-variables-1, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r geography-and-variables-1-hint-1, eval = FALSE}
get_acs(... = "...", 
        table = "...",
        ... = ...)
```

```{r geography-and-variables-1-test, include = FALSE}
get_acs(geography = "state", 
        table = "B01001",
        year = 2021)
```

### 

This code got all variables associated with table `B01001`, which covers sex broken down by age, from the 2017-2021 5-year ACS.

### Exercise 2

Rather than using the `variables` argument, you can supply a table name to the `table` parameter in `get_acs()`; this will return data for every variable in that table. 

Run `get_acs()` below setting `geography` to `"state"`, `table` to `"B01001"`, and `year` to `2021`

```{r geography-and-variables-2, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r geography-and-variables-2-hint-1, eval = FALSE}
get_acs(... = "...", 
        table = "...",
        ... = ...)
```

### 

This code got all variables associated with table `B01001`, which covers sex broken down by age, from the 2017-2021 5-year ACS.

### Exercise 3

What if we want to analyze a variable at the county level, but only in one state? We can use the `state` argument along with the geography argument.

### 

Run `get_acs()` below setting `geography` = `"county"`, `variables` to `"B19013_001"`, `state` to `"GA"`, and `year` to `2021`

```{r geography-and-variables-3, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r geography-and-variables-3-hint-1, eval = FALSE}
get_acs(... = "...", 
        variables = "...", 
        ... = "WI",
        year = ...)
```

### 

The example above returns average household income by county, but only in the state of Georgia. When using the `state` argument, you can use the state name (`"Georgia"`), the state postal code (`"GA"`), or the FIPS code (`"13"`).

### Exercise 4

Smaller geographies, such as Census tracts, can be subsetted as well by the `county` argument. 

Run `get_acs()` below setting `geography` = `"tract"`, `variables` to `"B19013_001"`, `state` to `"13"`, `county` to `"Gwinnett"`, and `year` to `2021`

```{r geography-and-variables-4, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r geography-and-variables-4-hint-1, eval = FALSE}
get_acs(geography = "...", 
        variables = "...", 
        state = "...", 
        ... = "Dane",
        ... = ...)
```

### 

The above code retrieves the average household income for every tract in Gwinnett County in Georgia. Subsetting is extremely useful, as it allows you to download only the data which is relevant to you.

### 

### Exercise 5

Let's switch gears. What if we don't know what a variable ID means; after all, they're just a bunch of letters and numbers.

Luckily, the **tidycensus** packages offers the `load_variables()` function. 

Run `load_variables()` setting `year` to `2021`, and `dataset` to `"acs5"`

```{r geography-and-variables-5, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r geography-and-variables-5-hint-1, eval = FALSE}
load_variables(year = ..., dataset = "...")
```

### 

Notice the "label" and "concept" columns, providing information about each variable. To learn more about `load_variables()` and its arguments, go to [Kyle Walkers introduction to tidycensus](https://walker-data.com/census-r/an-introduction-to-tidycensus.html#searching-for-variables-in-tidycensus)

### Exercise 6

Now we may know what a variable name means, but it's probably pretty annoying to type that every, single time you need it. So how can we rename it?

Run `get_acs()` below setting `geography` to `"county"`, `state` = `"Utah"`, `variables` to a vector containing `medinc` which is assigned to `"B19013_001"` and `medage` which is assigned to `"B01002_001"`, and `year` to `2021`.  

```{r geography-and-variables-6, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r geography-and-variables-6-hint-1, eval = FALSE}
get_acs(geography = "...",
        state = "...",
        ... = c(medinc = "...",
                      medage = "..."),
        year = ...)
```

### 

In the variable column, the values are "medinc" and "medage", not the complex ones we imported. 

### Exercise 7

Now, let's learn about `get_estimates()`.
This gets data from the **Population Estimates Program** (PEP). 

Run `get_estimates` below setting `geography` to `"county"`, `product` to `"components"`, `state` to `"GA"`, `county` to `"Chatham"`, and `year` to `2019`

```{r geography-and-variables-7, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r geography-and-variables-7-hint-1, eval = FALSE}
get_estimates(geography = "...",
              product = "...",
              ... = "...",
              county = "...",
              ... = ...)
```

### 

The code above requested all components of change population estimates for 2019 for Chatham County in Georgia. 

The confusing part of this function is probably the `product` argument, which essentially allows the user to access the indicators used to produce the intercensal population estimates. 

Learn more about `get_estimates()` and its arguments [here](https://walker-data.com/census-r/an-introduction-to-tidycensus.html#other-census-bureau-datasets-in-tidycensus)

### Exercise 8

`get_flows()` allows for analysis of immigrants, emigrants, and net migration for a given geography using data from a given 5-year ACS sample.

Run `get_flows()` below setting `geography` to `"county"`, `state` to `"HI"`, `county` to `"Honolulu"`, and `year` to `2019`

```{r geography-and-variables-8, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>


```{r geography-and-variables-8-hint-1, eval = FALSE}
get_flows(geography = "...",
          ... = "...",
          county = "...",
          ... = ...)
```

### 

This should return a tibble with 6 columns, one of them being "variable". We can see variables such as "MOVEDIN" and the corresponding estimate of 152 as the number of people who moved from Africa (look at the `FULL2_NAME` variable) to Honolulu in 2019

## Summary
### 

Here's your cheat-sheet of the core functions in tidycensus:

1. `get_decennial()`: requests data from the US Decennial Census APIs for 2000, 2010, and 2020.

2. `get_acs()`: requests data from the 1-year and 5-year American Community Survey samples. Data are available from the 1-year ACS back to 2005 and the 5-year ACS back to 2005-2009.

3. `get_estimates()`: an interface to the Population Estimates APIs. These datasets include yearly estimates of population characteristics by state, county, and metropolitan area, along with components of change demographic estimates like births, deaths, and migration rates.

4. `get_pums()`: which accesses data from the ACS Public Use Microdata Sample APIs. These samples include anonymized individual-level records from the ACS organized by household and are highly useful for many different social science analyses. 

5. `get_flows()`: an interface to the ACS Migration Flows APIs. Includes information on in- and out-flows from various geographies for the 5-year ACS samples, enabling origin-destination analyses.



```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
