---
title: Exploring US Census data with visualization
author: Satvika Upperla
tutorial:
  id: wrangling-census-data-with-tidyverse-tools
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: 'Tutorial for Chapter 4: Wrangling Census data with tidyverse tools'
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(tidycensus)
library(tidyverse)
library(knitr)
library(scales             )

knitr::opts_chunk$set(echo = FALSE)
options(tutorial.exercise.timelimit = 60, 
        tutorial.storage = "local") 

# We want the tutorial to run even if students do not have an internet
# connection, so we need to save all the downloaded data. Save the creation code
# to make replication/improvements easier.

# maine_income <- get_acs(
#  state = "Maine",
#  geography = "county",
#  variables = c(hhincome = "B19013_001"),
#  year = 2020
#) |>
#  mutate(NAME = str_remove(NAME, " County, Maine"))
# write_rds(maine_income, "data/main_income.rds")
 maine_income <- read_rds( "data/main_income.rds")


# years <- 2005:2019
#names(years) <- years

#deschutes_value <- map_dfr(years, ~{
 # get_acs(
 #   geography = "county",
 #   variables = "B25077_001",
  #  state = "OR",
  #  county = "Deschutes",
 #   year = .x,
 #   survey = "acs1"
 # )
#}, .id = "year")
#  write_rds(deschutes_value, "data/deschutes_value.rds")
 read_rds("data/deschutes_value.rds")
  
  

  
  
   # utah <- get_estimates(
 # geography = "state",
 #  state = "UT",
 # product = "characteristics",
 # breakdown = c("SEX", "AGEGROUP"),
 # breakdown_labels = TRUE,
 # year = 2019
#) 
 # write_rds(utah, "data/utah.rds")
  read_rds("data/utah.rds")
  
 # utah_filtered <- filter(utah, str_detect(AGEGROUP, "^Age"), 
 #                SEX != "Both sexes") |>
  #mutate(value = ifelse(SEX == "Male", -value, value))
  
#  write_rds(utah_filtered, "data/utah_filtered.rds")
  read_rds("data/utah_filtered.rds")
  
  
  
 # housing_val <- get_acs(
  #geography = "tract", 
  #variables = "B25077_001", 
  #state = "OR", 
#  county = c(
 #   "Multnomah", 
  #  "Clackamas", 
   # "Washington",
    #"Yamhill", 
  #  "Marion", 
  #  "Columbia"
  #),
  #year = 2020
#)
  
#write_rds(housing_val, "data/housing_val.rds")
read_rds("data/housing_val.rds")

#housing_val2 <- separate(
 # housing_val, 
  #NAME, 
  #into = c("tract", "county", "state"), 
  #sep = ", "
#)
#write_rds(housing_val2, "data/housing_val2.rds")
read_rds("data/housing_val2.rds")
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

This tutorial covers [Chapter 4: Exploring US Census data with visualization](https://walker-data.com/census-r/exploring-us-census-data-with-visualization.html#visualizing-margins-of-error) from [*Analyzing US Census Data: Methods, Maps, and Models in R*](https://walker-data.com/census-r/index.html) by Kyle Walker. You will learn about the [**ggplot**] and [**plotly**] package and mapping and using it to visualize moes (margin of error)

<!-- DK: Add note about skipping first two sections. -->

## Visualizing Margins of Error
### 

The [tidyverse](https://tidyverse.tidyverse.org/) is a collection of R packages that are designed to work together in common data wrangling, analysis, and visualization projects. Many of these R packages, maintained by RStudio, are among the most popular R packages worldwide.

### Exercise 1

Load in **tidyverse** package. 

```{r visualizing-margins-of-error-1, exercise = TRUE}

```

```{r visualizing-margins-of-error-1-hint-1, eval = FALSE}
library(...)
```

```{r visualizing-margins-of-error-1-test, include = FALSE}
library(tidyverse)
```

The **tidyverse** is not really a package itself, but, rather, loads several core packages.

### Exercise 2

We can access the US Census API via the **tidycensus** package. The **tidycensus** package utilizes an API key to access Census data. Load the **tidycensus** package into your R session.

Run `search()` in the Console to see the libraries that you've currently loaded. CP/CR.

```{r visualizing-margins-of-error-2}
question_text(NULL,
    answer(NULL, correct = TRUE),
    allow_retry = TRUE,
    try_again_button = "Edit Answer",
    incorrect = NULL,
    rows = 3)
```

### 

The output should include the string "package:tidycensus". Kyle Walker is also the author of the [**tidycensus**](https://walker-data.com/tidycensus/) package.

### Exercise 3

In the example below, we will compare the median household incomes of counties in the US state of Maine from the 2016-2020 ACS.

Use the function `get_decennial()` and passing in `state = "Maine"`, `geography = "county"`, `variables = c(totalpop = "P1_001N")`, and  `year = 2020` as arguments
 
```{r visualizing-margins-of-error-3, exercise = TRUE}

```

```{r visualizing-margins-of-error-3-hint-1, eval = FALSE}
get_decennial(
  state = ...,
  ... = "county",
  variables = c(... = "P1_001N"),
  year = 2020
) 
```

### Exercise 4

Pipe the previous code into the function  `arrange()` with `desc(value)` as an argument.
 
```{r visualizing-margins-of-error-4, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-margins-of-error-4-hint-1, eval = FALSE}
get_decennial(
  state = ...,
  ... = "county",
  variables = c(... = "P1_001N"),
  year = 2020
) |>
   arrange(...(value))
```

There are sixteen counties in Maine, ranging in population from a maximum of 303,069 to a minimum of 16,800. In turn, estimates for the counties with smaller population sizes are likely to be subject to a larger margin of error than those with larger baseline populations.

### Exercise 5

Comparing median household incomes of these sixteen counties illustrates this point. Let’s first obtain this data with tidycensus then clean up the NAME column with str_remove() to remove redundant information.
Delete the `arrange()` and anything within in it and replace it with the `mutate()` function and pass in `NAME = str_remove()` as the argument. withing `str_remove()` pass in `NAME, " County, Maine"` as an argument. 
 
```{r visualizing-margins-of-error-5, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-margins-of-error-5-hint-1, eval = FALSE}
get_decennial(
  state = ...,
  ... = "county",
  variables = c(... = "P1_001N"),
  year = 2020
) |>
   mutate(NAME = ....(..., " County, Maine"))
```

This will clean up the NAME column replacing something like "York County, Maine" with just "York" instead.

### Exercise 6

We have assigned the previous code to the variable `main_income`. Print `main_income` to ensure it works.

```{r visualizing-margins-of-error-6, exercise = TRUE}

```

```{r visualizing-margins-of-error-6-hint-1, eval = FALSE}
maine_income
```

```{r visualizing-margins-of-error-6-test, include = FALSE}
maine_income
```

### Exercise 7

Several county estimates on the chart are quite close to one another, which may mean that the ranking of counties is misleading given the margin of error around those estimates. Inspect the moe of these counties more. Pipe `main_income` to  `arrange(desc(moe))`

```{r visualizing-margins-of-error-7, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-margins-of-error-7-hint-1, eval = FALSE}
maine_income |>
  ...(desc(...))
```

```{r visualizing-margins-of-error-7-test, include = FALSE}
maine_income |>
  arrange(desc(moe))
```

### Exercise 8

Lets visualize the moe's of this data using ggplot. Pipe `main_income` to `ggplot()` setting `x` equal to `estimate` and `y` equal to `reoorder(NAME, estimate)` within `aes`.

```{r visualizing-margins-of-error-8, exercise = TRUE}

```

```{r visualizing-margins-of-error-8-hint-1, eval = FALSE}
ggplot(..., aes(x = ..., y = reorder(NAME, ...)))
```

```{r visualizing-margins-of-error-8-test, include = FALSE}
ggplot(maine_income, aes(x = estimate, y = reorder(NAME, estimate)))
```

### Exercise 9

Add `geom_point` setting `size` to `3` and `color` to `"darkgreen"` within the function

```{r visualizing-margins-of-error-9, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-margins-of-error-9-hint-1, eval = FALSE}
ggplot(..., aes(x = ..., y = reorder(NAME, ...))) +  
  geom_point(... = 3, color =  ...)
```

```{r visualizing-margins-of-error-9-test, include = FALSE}
ggplot(maine_income, aes(x = estimate, y = reorder(NAME, estimate))) +  
  geom_point(size = 3, color = "darkgreen")
```

### Exercise 10

```{r visualizing-margins-of-error-10, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-margins-of-error-10-hint-1, eval = FALSE}
ggplot(..., 
       aes(x = ..., 
           y = reorder(NAME, ...))) +  
  geom_point(... = 3, color =  ...)
```

```{r visualizing-margins-of-error-10-test, include = FALSE}
maine_income |> 
  ggplot(aes(x = estimate, 
             y = reorder(NAME, estimate))) +  
    geom_point(size = 3, color = "darkgreen")
```

A dot plot like the one above intended to visualize a ranking of county household incomes in Maine may be misleading. However, using visualization tools in ggplot2, we can visualize the uncertainty around each estimate, giving chart readers a sense of the uncertainty in the ranking.

### Exercise 11

 Lets add `geom_errorbar()` to visualize the margin of uncertainty. Pipe the previous code to `geom_errorbarh()` setting `xmin` equal to `estimate` minus `moe` and `xmax` to `estimate` plus `moe` within `aes()`.
 

```{r visualizing-margins-of-error-11, exercise = TRUE}

```

```{r visualizing-margins-of-error-11-hint-1, eval = FALSE}
ggplot(..., aes(x = ..., y = reorder(NAME, ...))) +  
  geom_point(... = 3, color =  ...) + 
  ...(aes(xmin = ... - moe, xmax = estimate + ...))
```

```{r visualizing-margins-of-error-11-test, include = FALSE}
ggplot(maine_income, aes(x = estimate, y = reorder(NAME, estimate))) +  
  geom_point(size = 3, color = "darkgreen") + 
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe))
```

Adding the horizontal error bars around each point gives us critical information to help us understand how our ranking of Maine counties by median household income.

### Exercise 12

Lets make the plot look nice using `labs()` and `theme_minimal()`. Pipe to `theme_minimal()` setting `base_size` to `12.5`. Using `labs()` add a `title`, `subtitle`, `x`, and `y`. Here is our version:


```{r}
ggplot(maine_income, aes(x = estimate, y = reorder(NAME, estimate))) +  
  geom_point(size = 3, color = "darkgreen") + 
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) + 
  theme_minimal(base_size = 12.5) + 
  labs(title = "Median household income", 
       subtitle = "Counties in Maine", 
       x = "2016-2020 ACS estimate", 
       y = "")
```

But your version does not need to be identical.

```{r visualizing-margins-of-error-12, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-margins-of-error-12-hint-1, eval = FALSE}
ggplot(..., aes(x = ..., y = reorder(NAME, ...))) +  
  geom_point(... = 3, color =  ...) + 
  ...(aes(xmin = ... - moe, xmax = estimate + ...)) + 
  theme_minimal(... = 12.5) + 
  labs(title = ..., 
      ... = "Counties in Maine", 
       x = "2016-2020 ACS estimate", 
       y = ...)
```

```{r visualizing-margins-of-error-12-test, include = FALSE}
ggplot(maine_income, aes(x = estimate, y = reorder(NAME, estimate))) +  
  geom_point(size = 3, color = "darkgreen") + 
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) + 
  theme_minimal(base_size = 12.5) + 
  labs(title = "Median household income", 
       subtitle = "Counties in Maine", 
       x = "2016-2020 ACS estimate", 
       y = "")
```

### Exercise 13

Let's make the numbers on the x=axis neater. Pipe the previous code into `scale_x_continuous()` setting `labels = label_dollar()` as the argument

```{r visualizing-margins-of-error-13, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-margins-of-error-13-hint-1, eval = FALSE}
ggplot(..., aes(x = ..., y = reorder(NAME, ...))) +  
  geom_point(... = 3, color =  ...) + 
  ...(aes(xmin = ... - moe, xmax = estimate + ...)) + 
  theme_minimal(... = 12.5) + 
  labs(title = ..., 
      ... = "Counties in Maine", 
       x = "2016-2020 ACS estimate", 
       y = ...) + 
  scale_x_continuous(labels = ...)
```

```{r visualizing-margins-of-error-13-test, include = FALSE}
ggplot(maine_income, aes(x = estimate, y = reorder(NAME, estimate))) +  
  geom_point(size = 3, color = "darkgreen") + 
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) + 
  theme_minimal(base_size = 12.5) + 
  labs(title = "Median household income", 
       subtitle = "Counties in Maine", 
       x = "2016-2020 ACS estimate", 
       y = "") + 
  scale_x_continuous(labels = label_dollar())
```


## Visualizing ACS Estimates Over Time
###
While the output table usefully represented the time series of educational attainment in Colorado counties, data visualization is also commonly used to illustrate change over time. Arguably the most common chart type chosen for time-series visualization is the line chart, which ggplot2 handles capably with the geom_line() function.

### Exercise 1

For an illustrative example, we’ll obtain 1-year ACS data from 2005 through 2019 on median home value for Deschutes County, Oregon, home to the city of Bend and large numbers of in-migrants in recent years from the Bay Area in California.

Assign the variable `years` to the years `2005` to `2019`. On another line assign the function `names()` with    `years` as its argument to the variable `years`.    
```{r visualizing-acs-estimates-over-1, exercise = TRUE}

```


```{r visualizing-acs-estimates-over-1-hint, eval = FALSE}
... <- 2005;2019
names(...) <- years
```

```{r visualizing-acs-estimates-over-1-test, include = FALSE}
years <- 2005:2019
names(years) <- years

```

### Exercise 2

Copy the previous code and start on a new line. Use the function `map_dfr()` setting `years` as an argument. the second argument will be a call to `get_acs()`. Use `~{}` and within that put the call to `get_acs()`. Set the arguments of `get_acs()` as;  geography = "county", variables = "B25077_001", state = "OR", county = "Deschutes", year = .x, and survey = "acs1". Set the third argument of `map_dfr()` to `.id = "year"`.

```{r visualizing-acs-estimates-over-2, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-acs-estimates-over-2-hint, eval = FALSE}
years <- 2005:2019
names(years) <- years

...(years, ~{
  get_acs(
    geography = ...,
    ... = "B25077_001",
    state = "OR",
    county = "Deschutes",
    year = ...,
    survey = "acs1"
  )
}, .id = ...)
```

```{r visualizing-acs-estimates-over-2-test, include = FALSE}
years <- 2005:2019
names(years) <- years

map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B25077_001",
    state = "OR",
    county = "Deschutes",
    year = .x,
    survey = "acs1"
  )
}, .id = "year")
```

`map_dfr()` is used to iterate over a named vector of years, creating a time-series dataset of median home value in `Deschutes County` since 2005, and we use the formula specification for anonymous functions so that `~ .x` translates to function(x) x.

### Exercise 3

 We have assigned the previous code to the variable `deschutes_value`. Print out 'deschutes_value' to verify the results. 
 
```{r visualizing-acs-estimates-over-3, exercise = TRUE}

```


```{r visualizing-acs-estimates-over-3-hint, eval = FALSE}
deschutes_value
```

```{r visualizing-acs-estimates-over-3-test, include = FALSE}
# SU: Had to comment it out because apparently it cant find it; deschutes_value

```

### Exercise 4

This information can be visualized with familiar ggplot2 syntax. `deschutes_value` is specified as the input dataset, with year mapped to the X-axis and estimate mapped to the y-axis. The argument `group = 1` is used to help ggplot2 understand how to connect the yearly data points with lines given that only one county is being visualized. `geom_line()` then draws the lines, and we layer points on top of the lines as well to highlight the actual ACS estimates.

Start a pipe, piping `deschutes_value` to `ggplot()` setting `x` equal to `year`, `y` equal to `estimate` and `group` equal to `1` within `aes()`. Then add `geom_line()` and `geom_point()`.
```{r visualizing-acs-estimates-over-4, exercise = TRUE}

```

```{r visualizing-acs-estimates-over-4-hint, eval = FALSE}
deschutes_value |> ...(aes(x = ..., y = estimate, ... = 1)) + 
  ... + 
  geom_point()
```

```{r visualizing-acs-estimates-over-4-test, include = FALSE}
# SU: Had to comment it out because apparently it cant find it; deschutes_value |> ggplot(aes(x = year, y = estimate, group = 1)) + 
  #geom_line() + 
  #geom_point()

```

The chart shows rising home values prior to the 2008 recession; a notable drop after the housing market crash; and rising values since 2011, reflecting increased demand from wealthy in-migrants from locations like the Bay Area. 

### Exercise 5

Let's build the margin of error information into the line chart like we did in the previous section. We’ll use the ggplot2 function geom_ribbon() to draw the margin of error interval around the line, helping represent uncertainty in the ACS estimates.

After `ggplot()` and before `geom_line()` add a pipe of `geom_ribbon()` setting `ymax = estimate + moe` and `ymin = estimate - moe` with `aes()`. Then add arguments of `fill = "navy"` and `alpha = 0.4` edit `geom_line()` and `geom_point()` to have `color = "navy"` as an argument. Add `size = 2` as the final argument to `geom_point()`.

```{r visualizing-acs-estimates-over-5, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-acs-estimates-over-5-hint, eval = FALSE}
deschutes_value |> ggplot(aes(x = ..., y = ..., group = 1))  + 
  ...(aes(ymax = ..., ymin = estimate - moe), 
              fill = ...,
              alpha = 0.4) + 
  geom_line(... = "navy") + 
  geom_point(color = "navy", ... = 2)
```

```{r visualizing-acs-estimates-over-5-test, include = FALSE}
# SU: Had to comment it out because apparently it cant find itdeschutes_value |> ggplot(aes(x = year, y = estimate, group = 1))  + 
  # geom_ribbon(aes(ymax = ..., ... = estimate - moe), 
  #             ... = "navy",
  #             alpha = ...) + 
  # geom_line(color = "navy") + 
  # geom_point(color = "navy", size = 2)

```

### Exercise 6

Pipe the previous code to `theme_minimal()` with the argument `base_size = 12`. Add another pipe to  `scale_y_continuous()`. Set `labels` to `label_dollar(scale = .001, suffix = "k) as an argument.

```{r visualizing-acs-estimates-over-6, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-acs-estimates-over-6-hint, eval = FALSE}
deschutes_value |> ggplot(aes(x = year, y = estimate, group = 1))  + 
  geom_ribbon(aes(ymax = estimate + moe, ymin = estimate - moe), 
              fill = "navy",
              alpha = 0.4) + 
  geom_line(... = "navy") + 
  geom_point(color = ..., size = 2) + 
  ...(base_size = 12) + 
  scale_y_continuous(... = label_dollar(... = .001, suffix = ...)) 
```

```{r visualizing-acs-estimates-over-6-test, include = FALSE}
# SU: Had to comment it out because apparently it cant find itdeschutes_value |> ggplot(aes(x = year, y = estimate, group = 1))  + 
  # geom_ribbon(aes(ymax = estimate + moe, ymin = estimate - moe), 
  #             fill = "navy",
  #             alpha = 0.4) + 
  # geom_line(color = "navy") + 
  # geom_point(color = "navy", size = 2) + 
  # theme_minimal(base_size = 12) + 
  # scale_y_continuous(labels = label_dollar(scale = .001, suffix = "k")) 

```

### Exercise 7

Add `labels` for the `title`, `caption`, `x`, and `y` using `labs()`. This is what our graph looks like:
```{r}
# SU: Had to comment it out because apparently it cant find itggplot(deschutes_value, aes(x = year, y = estimate, group = 1)) + 
  # geom_ribbon(aes(ymax = estimate + moe, ymin = estimate - moe), 
  #             fill = "navy",
  #             alpha = 0.4) + 
  # geom_line(color = "navy") + 
  # geom_point(color = "navy", size = 2) + 
  # theme_minimal(base_size = 12) + 
  # scale_y_continuous(labels = label_dollar(scale = .001, suffix = "k")) + 
  # labs(title = "Median home value in Deschutes County, OR",
  #      x = "Year",
  #      y = "ACS estimate",
  #      caption = "Shaded area represents margin of error around the ACS estimate")
```

Yours does not have to look exactly the same. 

```{r visualizing-acs-estimates-over-7, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r visualizing-acs-estimates-over-7-hint, eval = FALSE}
deschutes_value |> ggplot(aes(x = year, y = estimate, group = 1))  + 
  geom_ribbon(aes(ymax = estimate + moe, ymin = estimate - moe), 
              fill = "navy",
              alpha = 0.4) + 
  geom_line(... = "navy") + 
  geom_point(color = ..., size = 2) + 
  ...(base_size = 12) + 
  scale_y_continuous(... = label_dollar(... = .001, suffix = ...)) +
  labs(title = ...,
       x = ...,
       y = ...,
       caption = ...)
```

```{r visualizing-acs-estimates-over-7-test, include = FALSE}
# SU: Had to comment it out because apparently it cant find itggplot(deschutes_value, aes(x = year, y = estimate, group = 1)) + 
  # geom_ribbon(aes(ymax = estimate + moe, ymin = estimate - moe), 
  #             fill = "navy",
  #             alpha = 0.4) + 
  # geom_line(color = "navy") + 
  # geom_point(color = "navy", size = 2) + 
  # theme_minimal(base_size = 12) + 
  # scale_y_continuous(labels = label_dollar(scale = .001, suffix = "k")) + 
  # labs(title = "Median home value in Deschutes County, OR",
  #      x = "Year",
  #      y = "ACS estimate",
  #      caption = "Shaded area represents margin of error around the ACS estimate")
```




```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
