---
title: 'Wrangling Census data with tidyverse tools'
author: David Kane
tutorial:
  id: wrangling-census-data-with-tidyverse-tools
output:
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: 'Tutorial for Chapter 3: Wrangling Census data with tidyverse tools'
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(tidycensus)
library(tidyverse)


knitr::opts_chunk$set(echo = FALSE)
options(tutorial.exercise.timelimit = 60, 
        tutorial.storage = "local") 

# We want the tutorial to run even if students do not have an internet
# connection, so we need to save all the downloaded data. Save the creation code
# to make replication/improvements easier.

# The tidyverse

# median_age <- get_acs(
#   geography = "county",
#   variables = "B01002_001",
#   year = 2020
# )
# write_rds(median_age, "data/median_age.rds")

median_age <- read_rds("data/median_age.rds")

race_vars <- c(
  White = "B03002_003",
  Black = "B03002_004",
  Native = "B03002_005",
  Asian = "B03002_006",
  HIPI = "B03002_007",
  Hispanic = "B03002_012"
)

# az_race <- get_acs(
#   geography = "county",
#   state = "AZ",
#   variables = race_vars,
#   summary_var = "B03002_001",
#   year = 2020
# ) 
# write_rds(az_race, "data/az_race.rds")

az_race <- read_rds("data/az_race.rds")

az_race_percent <- az_race %>%
  mutate(percent = 100 * (estimate / summary_est)) %>%
  select(NAME, variable, percent)

# Groupwise

# mn_hh_income <- get_acs(
#   geography = "county",
#   table = "B19001",
#   state = "MN",
#   year = 2016
# )
# write_rds(mn_hh_income, "data/mn_hh_income.rds")

mn_hh_income <- read_rds("data/mn_hh_income.rds")

mn_hh_income_recode <- mn_hh_income %>%
  filter(variable != "B19001_001") %>%
  mutate(incgroup = case_when(
    variable < "B19001_008" ~ "below35k", 
    variable < "B19001_013" ~ "bw35kand75k", 
    TRUE ~ "above75k"
  )) 

# Comparing ACS estimates over time


# oglala_lakota_age <- get_acs(
#   geography = "county",
#   state = "SD",
#   county = "Oglala Lakota",
#   table = "B01001",
#   year = 2020)
# write_rds(oglala_lakota_age, "data/oglala_lakota_age.rds")

oglala_lakota_age <- read_rds("data/oglala_lakota_age.rds")

college_vars <- c("B15002_015",
                  "B15002_016",
                  "B15002_017",
                  "B15002_018",
                  "B15002_032",
                  "B15002_033",
                  "B15002_034",
                  "B15002_035")

years <- 2010:2019
names(years) <- years

# college_by_year <- map_dfr(years, ~{
#   get_acs(
#     geography = "county",
#     variables = college_vars,
#     state = "CO",
#     summary_var = "B15002_001",
#     survey = "acs1",
#     year = .x
#   )
# }, .id = "year")
# write_rds(college_by_year, "data/college_by_year.rds")  

college_by_year <- read_rds("data/college_by_year.rds")
  
percent_college_by_year <- college_by_year %>%
  group_by(NAME, year) %>%
  summarize(numerator = sum(estimate),
            denominator = first(summary_est)) %>%
  mutate(pct_college = 100 * (numerator / denominator)) %>%
  pivot_wider(id_cols = NAME,
              names_from = year,
              values_from = pct_college)
  
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
### 

This tutorial covers [Chapter 3: Wrangling Census data with tidyverse tools](https://walker-data.com/census-r/wrangling-census-data-with-tidyverse-tools.html) from [*Analyzing US Census Data: Methods, Maps, and Models in R*](https://walker-data.com/census-r/index.html) by Kyle Walker. You will learn about the [**tidycensus**](https://walker-data.com/tidycensus/index.html) package and some of its key functions, including [`get_decennial()`](https://walker-data.com/tidycensus/articles/basic-usage.html) and [`get_acs()`](https://walker-data.com/tidycensus/articles/basic-usage.html). We discuss margins of error (MOEs) in the American Community Survey and how to wrangle and interpret MOEs appropriately.

## The tidyverse
### 

<!-- Covers sections 3.1 and 3.2. -->

The [tidyverse](https://tidyverse.tidyverse.org/) is a collection of R packages that are designed to work together in common data wrangling, analysis, and visualization projects. Many of these R packages, maintained by RStudio, are among the most popular R packages worldwide.

### Exercise 1

Start your code by loading in the **tidyverse** and **tidycensus** packages. 

```{r the-tidyverse-1, exercise = TRUE}

```

```{r the-tidyverse-1-hint-1, eval = FALSE}
library(...)
library(...)
```

```{r the-tidyverse-1-test, include = FALSE}
library(tidyverse)
library(tidycensus)
```

The tidyverse is a collection of R packages that are designed to work together in common data wrangling, analysis, and visualization projects. The **tidyverse** specifically a package itself, but rather loads several core packages within the tidyverse.

### Exercise 2

For a first example, let’s request data on median age from the 2016-2020 ACS with `get_acs()` for all counties in the United States. Run `get_acs()` with `geography` = `"county"`, `variables` = `"B01002_001"`, and the `year` = `2020`.

```{r the-tidyverse-2, exercise = TRUE}

```

```{r the-tidyverse-2-hint-1, eval = FALSE}
get_acs(
  geography = "...",
  variables = "B01002_001",
  year = ...
)
```

### 

The default method for printing data used by the **tibble** package shows the first 10 rows of the dataset, which in this case prints counties in Alabama. On the right hand side of the tibble you can see both moe and estimate, both of which we will talk about later in the chapter. 


### Exercise 3

Set `median_age` to the result of your previous call to `get_acs()`.


```{r the-tidyverse-3, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r the-tidyverse-3-hint-1, eval = FALSE}
... <- get_acs(
  geography = "county",
  variables = "B01002_001",
  year = 2020
)
```

### 

We almost always assign the results of a function call which retrieves data from a source like the Census to a permanent object, `median_age` in this case. Strictly speaking, we don't need to do this. We could just start a pipe with this function call. But, first, this is sometimes annoyingly slow and, second, we don't want to be abusive and query an online source every time we make a tiny change.

### Exercise 4

Write `median_age` and press "Run Code".


```{r the-tidyverse-4, exercise = TRUE}

```

```{r the-tidyverse-4-hint-1, eval = FALSE}
median_age
```

```{r the-tidyverse-4-test, include = FALSE}
median_age
```

### 

Note that the key variable is called `estimate` rather than `age` or `median_age`. The default behavior of `get_acs()` is for the variable of interest to be labelled `estimate` since, in fact, it is an estimate. 



### Exercise 5

`arrange()` sorts a dataset by values in one or more columns and returns the sorted result. Pipe `median_age` to the `arrange()` with the argument `estimate`.

```{r the-tidyverse-5, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r the-tidyverse-5-hint-1, eval = FALSE}
median_age |> 
  ...(estimate)
```

```{r the-tidyverse-5-test, include = FALSE}
median_age |> 
  arrange(estimate)
```

### 

Per the 2016-2020 ACS, the youngest county is De Baca County, New Mexico. Two of the five youngest “counties” in the United States are independent cities in Virginia, which are treated as county-equivalents. 

We would normally do this analysis by justing piping `median_age` into `arrange(estimate)`, producing the same answer as here.


### Exercise 6

New let's try to find the oldest counties in the US by median age. Change `estimate` to `desc(estimate)` within `arrange()` in the previous code.

```{r the-tidyverse-6, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r the-tidyverse-6-hint-1, eval = FALSE}
median_age |> 
  arrange(...(estimate))
```

```{r the-tidyverse-6-test, include = FALSE}
median_age |> 
  arrange(desc(estimate))
```

### 

The oldest county in the United States is Sumter County, Florida. Sumter County is home to The Villages, a Census-designated place that includes a large age-restricted community also called [The Villages](https://www.thevillages.com/). 



### Exercise 7

How many counties in the US have a median age of 50 or older? Pipe `median_age` to `filter()` with the argument `estimate >= 50`. 


```{r the-tidyverse-7, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r the-tidyverse-7-hint-1, eval = FALSE}
median_age |> 
  filter(... >= 50)
```

```{r the-tidyverse-7-test, include = FALSE}
median_age |> 
  filter(estimate >= 50)
```

### 

The `filter()` function queries a data set for rows where a given condition evaluates to TRUE, and retains those rows only. In this case, the data set is filtering our `median_age` data set for counties that have a median age of 50 or older. 


### Exercise 8

What if we want to separate the columns into county and state? The best tool for that is the `separate_wider_delim()` function. Pipe `median_age` to `separate_wider_delim()` with arguments `col = name`, `into = c("county","state")` and `sep = ", "`. (Note that it is `", "` with a space, not `","` without one.)

```{r the-tidyverse-8, exercise = TRUE}

```

```{r the-tidyverse-8-hint-1, eval = FALSE}
median_age |> 
  separate(... = NAME,
  into = ...,
  ... = ", ")
```

```{r the-tidyverse-8-test, include = FALSE}
median_age |> 
  separate(col = NAME,
  into = c("county", "state"),
  sep = ", ")
```

### 

`separate()` is an older **dplyr** function for accomplishing this same task. It has been superseded by [`separate_wider_position()`](https://tidyr.tidyverse.org/reference/separate_wider_delim.html) and [`separate_wider_delim()`](https://tidyr.tidyverse.org/reference/separate_wider_delim.html). 


### Exercise 9

Data in Census and ACS tables, as in the example above, are frequently comprised of variables that individually constitute sub-categories such as the numbers of households in different household income bands. One limitation of the approach above, however, is that the data and the resulting analysis return estimated counts, which are difficult to compare across geographies. Press "Run Code".


```{r the-tidyverse-9, exercise = TRUE}
race_vars <- c(
  White = "B03002_003",
  Black = "B03002_004",
  Native = "B03002_005",
  Asian = "B03002_006",
  HIPI = "B03002_007",
  Hispanic = "B03002_012"
)
```

```{r the-tidyverse-9-test, include = FALSE}
race_vars <- c(
  White = "B03002_003",
  Black = "B03002_004",
  Native = "B03002_005",
  Asian = "B03002_006",
  HIPI = "B03002_007",
  Hispanic = "B03002_012"
)
```

### 

We will make use of this variable in our next download request. 

### Exercise 10

Call `get_acs()` with `geography = "county"`, `state = "AZ"`, `variables = race_vars`,  and `year = 2020`. This is similar to previous calls we have made to `get_acs()` except, instead of using a single variable code, we are providing a vector of codes.  


```{r the-tidyverse-10, exercise = TRUE}

```

```{r the-tidyverse-10-hint-1, eval = FALSE}
get_acs(
  geography = "county",
  state = "AZ",
  variables = race_vars,
  year = 2020
) 
```

### 

Note the wide range in raw numbers. Some categories are 100 or even 1,000 times more common than others. We often want to "normalize" these values so that everything is on a more similar scale.



### Exercise 11

A solution to this issue might involve normalizing the estimated count data by dividing it by the overall population from which the sub-group is derived. Appropriate denominators for ACS tables are frequently found in the tables themselves as variables.

Use the same code as in the last Exercise, but add `summary_var = "B03002_001"` as an argument to the call to `get_acs()`.


```{r the-tidyverse-11, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r the-tidyverse-11-hint-1, eval = FALSE}
get_acs(
  geography = "county",
  state = "AZ",
  variables = race_vars,
  summary_var = "B03002_001",
  year = 2020
) 
```

### 

Adding a variable ID to the `summary_var` parameter in `get_acs()`  will create two new columns: `summary_est` and `summary_moe`, representing the summary variable’s estimate and margin of error. In this case, the summary variable is the total population for each county. 

### Exercise 12

In the background, we ran this same download and assigned the result to a new object: `az_race`. Type `az_race` and hit "Run Code".

```{r the-tidyverse-12, exercise = TRUE}

```

```{r the-tidyverse-12-hint-1, eval = FALSE}
az_race
```

### 

Look closely as the results for Apache County. Each of the variables which we requested gets its own row. Instead of just having one row for the county, we have 6 rows, one for each race. The `summary_est` is the total population of the county. (Note that, for technical reasons, the sum of the estimated populations for each race may not exactly equal the estimated population for the whole county.)

<!-- DK: Really ought to understand and explain why `sum(az_race$estimate[1:6])` does not equal `az_race$summary_est[1]`. -->

### Exercise 13

Pipe `az_race` to `mutate(percent = 100 * (estimate / summary_est))` and continue the pipe with `select(NAME, variable, percent)`.

```{r the-tidyverse-13, exercise = TRUE}

```


```{r the-tidyverse-13-hint-1, eval = FALSE}
az_race %>%
  ...(percent = 100 * (estimate / ...)) %>%
  select(..., variable, percent)
```

```{r the-tidyverse-13-test, include = FALSE}
az_race %>%
  mutate(percent = 100 * (estimate / summary_est)) %>%
  select(NAME, variable, percent)
```

### 

This is the typical workflow. First, we figure out what variables we need and download them using a function like `get_acs()` or `get_decennial()`. We save the result to a permanent object like `az_race`. Then, we build pipes starting from that saved object.






## Comparing ACS estimates over time
### 

A common task when working with Census data is to examine demographic change over time. Data from the Census API - and consequently tidycensus - only go back to the 2000 Decennial Census.

### Exercise 1

Before engaging in any sort of time series analysis of Census data, analysts need to account for potential problems that can emerge when using Census data longitudinally. One major issue that can emerge is geography changes over time.
 

Use `get_acts()` to create the dataset `oglala_lakota_age`. Within `get_acts()`, set `geography` = `"county"`, `state` = `"SD"`, `county` = `"Oglala Lakota"`, `table` =  `"B01001"` and `year` = `2020`. 

```{r comparing-acs-estimates-over-t-1, exercise = TRUE}

```

```{r comparing-acs-estimates-over-t-1-hint-1, eval = FALSE}
oglala_lakota_age <- ...(
  geography = "...",
  state = "...",
  ... = "Oglala Lakota",
  table = "...",
  year = ...
)
```

### 

The `get_acs()` function allows us obtain data from the American Community Survey. More information on this function can be found [here](https://www.rdocumentation.org/packages/tidycensus/versions/1.5/topics/get_acs).  



### Exercise 2

Print out the `oglala_lakota_age` tibble which you just created.

```{r comparing-acs-estimates-over-t-2, exercise = TRUE}

```

```{r comparing-acs-estimates-over-t-2-hint-1, eval = FALSE}
oglala_lakota_age
```

```{r comparing-acs-estimates-over-t-2-test, include = FALSE}
oglala_lakota_age
```

### 

This is a 2016-2020 age table for Oglala Lakota County, SD. This graph allows us to see the age composition of the Oglola Lakota County through this time period


### Exercise 3

Using the same code as in the first exercise in this section, change the `year` = `2010`. You should get an error. 

```{r comparing-acs-estimates-over-t-3, exercise = TRUE}

```

```{r comparing-acs-estimates-over-t-3-hint-1, eval = FALSE}
oglala_lakota_age <- get_acs(
  geography = "county",
  state = "SD",
  county = "Oglala Lakota",
  table = "B01001",
  year = ...
)
```

### 

We get an error because ten years ago, this county was named "Shannon". 

### Exercise 4

Using the same code as above, change the `county` name to `"Shannon"`. 

```{r comparing-acs-estimates-over-t-4, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-4-hint-1, eval = FALSE}
oglala_lakota_age <- get_acs(
  geography = "county",
  state = "SD",
  county = "...",
  table = "B01001",
  year = 2010
)
```

```{r comparing-acs-estimates-over-t-4-test, include = FALSE}
oglala_lakota_age <- get_acs(
  geography = "county",
  state = "SD",
  county = "Shannon",
  table = "B01001",
  year = 2010
)
```

### 

When looking at the two datasets, you may notice that the `GEOID` column changed from `46113` in 2010 to `46102`to 2020. This is because when a geographic entity changes their name, the Census Bureau assigns it a new `GEOID`, meaning that analysts need to take care when dealing with those changes. 


### Exercise 5

 Let’s say that we are interested in analyzing the percentage of residents age 25 and up with a 4-year college degree for counties in Colorado from the 2019 1-year ACS. 
 
 Set the dataset `co_college` to the function `get_acs()`. Inside the function, set `geography` = `"county"`, `variables` = `"DP02_0068P"`, `state` = `"CO"`, `survey` = `"acs1"` and `year` = `2019`.  
 

```{r comparing-acs-estimates-over-t-5, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-5-hint-1, eval = FALSE}
... <- get_acs(
  ... = "county",
  variables = "...",
  ... = "CO",
  survey = "...",
  year = ...
)
```

```{r comparing-acs-estimates-over-t-5-test, include = FALSE}
co_college19 <- get_acs(
  geography = "county",
  variables = "DP02_0068P",
  state = "CO",
  survey = "acs1",
  year = 2019
)
```

### 

Notice how Boulder County, home to the University of Colorado, has a very high percentage of its population with a 4-year degree or higher.

### Exercise 6

Copying the code above, lets change the `year` to `2018`. 

```{r comparing-acs-estimates-over-t-6, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-6-hint-1, eval = FALSE}
co_college19 <- get_acs(
  geography = "county",
  variables = "DP02_0068P",
  state = "CO",
  survey = "acs1",
  year = ...
)
```

```{r comparing-acs-estimates-over-t-6-test, include = FALSE}
co_college19 <- get_acs(
  geography = "county",
  variables = "DP02_0068P",
  state = "CO",
  survey = "acs1",
  year = 2018
)
```

### 

Why are the values completely changed and different? This is because variable ID's change every year and as a result should not be used for time-series analysis. These variable ID's actually represent the civilian population age 18 and up, and have nothing to do with educational attainment.

### Exercise 7

If our data is compromised from geography changes and variable ID changes, what is the best way to make a time-series analysis? The safest way to compare data is through Comparison Profile Tables!

Let's get data from the ACS Comparison Profile on inflation-adjusted median household income for counties and county-equivalents in Alaska.

Set the dataset `ak_income_compare` to the function `get_acs()`. Inside the function, set `geography` = `"county"`, `variables` to a vector of `income15 = "CP03_2015_062"` and `income20 = "CP03_2020_062"`, `state` = `"AK"`, and `year` = `2020`.

```{r comparing-acs-estimates-over-t-7, exercise = TRUE}

```

```{r comparing-acs-estimates-over-t-7-hint-1, eval = FALSE}
... <- get_acs(
  ... = "county",
  variables = c(
    income15 = "...",
    income20 = "..."
  ),
  ... = "AK",
  year = ...
)
```

```{r comparing-acs-estimates-over-t-7-test, include = FALSE}
ak_income_compare <- get_acs(
  geography = "county",
  variables = c(
    income15 = "CP03_2015_062",
    income20 = "CP03_2020_062"
  ),
  state = "AK",
  year = 2020
)
```

### 

These tables are available for both the 1-year and 5-year ACS, and allow for comparison of demographic indicators over the past five years for a given year. For the 2016-2020 ACS, the “comparison year” is 2015, representing the closest non-overlapping 5-year dataset, which in this case is 2011-2015.

### Exercise 8

Let’s re-engineer the analysis above on educational attainment in Colorado counties, which below will be computed for a time series from 2010 to 2019. We only need a few variables (representing estimates of populations age 25+ who have finished a 4-year degree or graduate degrees, by sex), we’ll request those variables directly rather than the entire B15002 table. 

Set `college_wars` equal to a vector of `"B15002_015"`, `"B15002_016"`, `"B15002_017"`, `"B15002_018"`, `"B15002_032"`, `"B15002_033"`, `"B15002_034"`, and `"B15002_035"`. 



```{r comparing-acs-estimates-over-t-8, exercise = TRUE}

```

```{r comparing-acs-estimates-over-t-8-hint-1, eval = FALSE}
college_vars <- c("...",
                  "B15002_016",
                  "B15002_017",
                  "...",
                  "...",
                  "B15002_033",
                  "...",
                  "B15002_035")
```

```{r comparing-acs-estimates-over-t-8-test, include = FALSE}
college_vars <- c("B15002_015",
                  "B15002_016",
                  "B15002_017",
                  "B15002_018",
                  "B15002_032",
                  "B15002_033",
                  "B15002_034",
                  "B15002_035")
```

### 

If we want data on college career holders from 2010 to 2019, we will have a lot of `get_acs()` functions. To reduce the unnecessary work, we will use iterations to avoid repetitive coding. This is similar to the loop operators `for` an `while` 


### Exercise 9

The **purr** package has a variety of iteration functions such as `map_*()`. The function `map_dfr()` is especially useful in tidycensus, as it iterates over an input and applies it to a function or process defined by the user, then row-binds the result into a single data frame. 

Set `years` to `2010:2019`. On a separate line type the function `names()` with the argument as `years`. Set this function to `years`. 



```{r comparing-acs-estimates-over-t-9, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-9-hint-1, eval = FALSE}
years <- ....
names(..) <- years
```

```{r comparing-acs-estimates-over-t-9-test, include = FALSE}
years <- 2010:2019
names(years) <- years
```

### 

The `names()` function allows us to get or set the names of an object. More information on this function can be found [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/names)

### Exercise 10

Copy the code above. Then set `college_by_year` to the function `map_dfr()` with the first argument as `years`. Inside the `map_dfr()` function, set the second argument to `~{ getacs() }`. It is fine to have an error.  

```{r comparing-acs-estimates-over-t-10, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-10-hint-1, eval = FALSE}

years <- 2010:2019
names(years) <- years

college_by_year <- ...(..., ~{
  get_acs() })
```

```{r comparing-acs-estimates-over-t-10-test, include = FALSE, eval = FALSE}
years <- 2010:2019
names(years) <- years

college_by_year <- map_dfr(years, ~{
  get_acs() })
```

### 

The tilde (`~`) operator and the curly brackets (`{}`) means that the code will run once for each element of `years`. 

### Exercise 11

Copy the code above. Within the `get_acs()` function, set `geography` = `"county"`, `variables` = `collge_vars`, `state` = `"CO"`, `summary_var` = `"B15002_001"`, `survey` = `"acs1"` and `year` = `.x`. Within the `map_dfr()` function, set the third argument `.id` equal to `"year"`

```{r comparing-acs-estimates-over-t-11, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-11-hint-1, eval = FALSE}
years <- 2010:2019
names(years) <- years

college_by_year <- map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = ...,
    ... = "CO",
    summary_var = "...",
    ... = "acs1",
    year = .x
  )
}, .id = "...")
```

```{r comparing-acs-estimates-over-t-11-test, include = FALSE}
years <- 2010:2019
names(years) <- years

college_by_year <- map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = college_vars,
    state = "CO",
    summary_var = "B15002_001",
    survey = "acs1",
    year = .x
  )
}, .id = "year")
```

### 

Let's go over the code so that you can understand. A numeric vector of `years` is defined with the syntax 2010:2019. This will create a vector of `years` at 1-year intervals. 

Then the function map_dfr() takes three arguments. The first argument, `years` is the object that the function will iterate over. The second argument uses the formula to create data for each element of `years` and then outputs the data into a single data frame. The third argument makes a new column that will output the data for `year`. 

### Exercise 12

Lets look at the results! 

Pipe `college_by_year` using `%>%` to the function `arrange()`. Within `arrange()`, set the first argument as `NAME`, the second as `variable` and the third as `year`. 

```{r comparing-acs-estimates-over-t-12, exercise = TRUE}

```

```{r comparing-acs-estimates-over-t-12-hint-1, eval = FALSE}
... %>% 
  arrange(NAME, ..., ...)
```

```{r comparing-acs-estimates-over-t-12-test, include = FALSE}
college_by_year %>% 
  arrange(NAME, variable, year)
```

### 

The result is a long-form data set that contains a time series of each requested ACS variable for each county in Colorado that is available in the 1-year ACS. 

### Exercise 13

Let's calculate the percentage of the population age 25 and up with a 4-year college degree using our data from before.  

Set `percent_college_by_year` to `college_by_year`. Then pipe `college_by_year` using `%>%` to the `group_by()` function with the first argument as `NAME` and the second argument as `year`. 
```{r comparing-acs-estimates-over-t-13, exercise = TRUE}

```

```{r comparing-acs-over-t-12-hint, eval = FALSE }
percent_college_by_year <- ... %>%
  group_by(..., year)
```

```{r comparing-acs-estimates-over-t-13-test, include = FALSE}
percent_college_by_year
```

### 

### Exercise 14

Copy the code above. Continue using the pipe `%>%` to the function `summarize()`. For the first argument of `summarize()`, set `numerator` equal to `sum(estimate)`. For the second argument of `summarize()`, set `denominator` to `first(summary_est)`. 

```{r comparing-acs-estimates-over-t-14, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-14-hint-1, eval = FALSE}
... <- college_by_year %>%
  group_by(..., year) %>%
  summarize(... = sum(estimate),
            ... = first(...))
```

```{r comparing-acs-estimates-over-t-14-test, include = FALSE}
percent_college_by_year
```

### 

The function `summarize()` creates a new data frame. More information can be found [here](https://dplyr.tidyverse.org/reference/summarise.html)

### Exercise 15

Copy the code above. Continue the pipe `%>%` to the function `mutate()`. Within the function, set `pct_college` = `100` times `(numerator/denominator)`. 

```{r comparing-acs-estimates-over-t-15, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-15-hint-1, eval = FALSE}
percent_college_by_year <- college_by_year %>%
  group_by(NAME, year) %>%
  summarize(numerator = sum(estimate),
            denominator = first(summary_est)) %>%
  mutate(... = ... * (numerator / ...)) 

```

```{r comparing-acs-estimates-over-t-15-test, include = FALSE}
percent_college_by_year
```

### 

The `mutate()` function allows us the create, modify and delete columns. More information about this function can be found [here](https://dplyr.tidyverse.org/reference/mutate.html)

### Exercise 16

Copy the code above. Continue the pipe `%>%` to the function `pivot_wider()`. Within the function, set `id_cols` = `NAME`, `names_from` = `year`, and `values_from` = `pct_college`. 

```{r comparing-acs-estimates-over-t-16, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r comparing-acs-estimates-over-t-16-hint-1, eval = FALSE}
percent_college_by_year <- college_by_year %>%
  group_by(NAME, year) %>%
  summarize(numerator = sum(estimate),
            denominator = first(summary_est)) %>%
  mutate(pct_college = 100 * (numerator / denominator)) %>%
  ...(... = NAME,
              names_from = ...,
              ... = pct_college)
```

```{r comparing-acs-estimates-over-t-16-test, include = FALSE}
percent_college_by_year
```

### 

The code below outlines a `group_by() %>% summarize()` workflow for calculating the percentage of the population age 25 and up with a 4-year college degree, then uses the `pivot_wider()` function from the tidyr package to spread the years across the columns for tabular data display. If you want to know more information about the `pivot_wider()` function click [here](https://tidyr.tidyverse.org/reference/pivot_wider.html)

## Handling margins of error with tidycensus
### 

The American Community Survey is based on a sample with estimates characterized by margins of error, and by default MOE is set to a 90 confidence interval. This can be translated roughly as “we are 90 percent sure that the true value falls within a range defined by the estimate plus or minus the margin of error.” 

### Exercise 1

Lets look at data on median household income by county in Rhode Island using the default `moe_level` of 90. 

Type the function `get_acs()`. Within this function, set `geography` = `"county"`, `state` = `"Rhode Island"`, `variables` = `"B19013_001"`, and `year` = `2020`. 

```{r handling-margins-of-error-with-1, exercise = TRUE}

```

```{r handling-margins-of-error-with-1-hint-1, eval = FALSE}
get_acs(
  geography = ...,
  ... = "Rhode Island",
  ... = ...,
  year = ...
)
```

```{r handling-margins-of-error-with-1-test, include = FALSE}
get_acs(
  geography = "county",
  state = "Rhode Island",
  variables = "B19013_001",
  year = 2020
)
```

### 

The data can be translated to saying, "We are 90% confident that the median household income in the county of Rhode Island falls within the range of the `estimate` plus or minus the `moe`." For example, our estimate of the average household income in Bristol County, Rhode Island is $85,413 with a 90% confidence interval of +/- $6,122.

### Exercise 2

Rather than using the default 90% margin of error, lets change the argument using the `get_acs()` function. Side note, the `moe_level` argument can only be either 90, 95 or 99.

Copy the code above and add an argument `moe_level` = `99`. 

```{r handling-margins-of-error-with-2, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r handling-margins-of-error-with-2-hint-1, eval = FALSE}
get_acs(
  geography = "county",
  state = "Rhode Island",
  variables = "B19013_001",
  year = 2020,
  moe_level = ...
)
```

```{r handling-margins-of-error-with-2-test, include = FALSE}
get_acs(
  geography = "county",
  state = "Rhode Island",
  variables = "B19013_001",
  year = 2020,
  moe_level = 99
)
```

### 

Look at the data and notice how the values for the `moe` column has changed. The stricter margin of error increases the size of the confidence interval. In simpler terms, if you want to be more confident (i.e., set a higher `moe_level`) in the estimate, then your margin of error will increase. 

### Exercise 3

For small geographies or small populations, margins of error can get quite large, in some cases exceeding their corresponding estimates. Let's examine data on age groups by sex for the population age 65 and older for Census tracts in Salt Lake County, Utah.

Set `vars` to the function `paste0()`. Within the function, set the first argument as `"B01001_0"` and the second arguement as a vector of `20:25` and `44:49`. On the next line print out `vars`. 

```{r handling-margins-of-error-with-3, exercise = TRUE}

```

```{r handling-margins-of-error-with-3-hint-1, eval = FALSE}
... <- paste0(..., c(20:25, ...))

vars
```

```{r handling-margins-of-error-with-3-test, include = FALSE}
vars <- paste0("B01001_0", c(20:25, 44:49))

vars
```

### 

Let's look at this code. First we are evaluating the `20:25` and `44:49` expressions to generate a vector of numbers for suffixes of variable IDs. Next we are using the `c()` function to combine the two vectors into a single vector. Lastly the `paste0()` function concatenates the prefix `"B01001_0"` with the variable IDs we just created. 

### Exercise 4

The resulting object `vars` allows us to request variables in a call to `get_acs()`. 

Set `salt_lake` to the function `get_acs()`. Within the function, set `geography` = `"tract"`, `variables` = `vars`, state = `"Utah"`, `county` = `"Salt Lake"`, and `year` = `2020`. 

```{r handling-margins-of-error-with-4, exercise = TRUE}

```

```{r handling-margins-of-error-with-4-hint-1, eval = FALSE}
... <- get_acs(
  ... = "tract",
  ... = vars,
  state = ...,
  county = ...,
  ... = 2020
)
```

```{r handling-margins-of-error-with-4-test, include = FALSE}
salt_lake <- get_acs(
  geography = "tract",
  variables = vars,
  state = "Utah",
  county = "Salt Lake",
  year = 2020
)
```

### 

The `paste0()` function we used for the `vars` object allows us to concatenate vectors after converting to character. More information about the function and its arguments can be found [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/paste.html)

### Exercise 5

Now we should examine our data using the `filter()` function. 

Set `example_tract` to the `salt_lake` object we used above. Now pipe the `salt_lake` object using `%>%` to the function `filter()`. Within `filter()`, set `GEOID` equals to `"49035100100"`. On a separate line, pipe `example_tract` to the `select()` function. Within the function, type `-NAME`.  

```{r handling-margins-of-error-with-5, exercise = TRUE}

```

```{r handling-margins-of-error-with-5-hint-1, eval = FALSE}
example_tract <- ... %>%
  filter(GEOID == ...)

example_tract %>% 
  select(...)
```

```{r handling-margins-of-error-with-5-test, include = FALSE}
example_tract <- salt_lake %>%
  filter(GEOID == "49035100100")

example_tract %>% 
  select(-NAME)
```

### 

In many cases, the ma
rgins of error exceed their corresponding estimates. For example, for the variable `B01001_0025` it suggests that or the male population age 85 and up, there are anywhere between 0 and 45 people in that Census tract. This can make ACS data for small geographies problematic for planning and analysis purposes.


### Exercise 6

A potential solution to large margins of error for small estimates in the ACS is to aggregate data upwards until a satisfactory margin of error to estimate ratio is reached.

The `moe_prop()` function calculates a margin of error for a derived proportion. Type the function, then set the arguements to `25`, `100`, `5`, and `3`. 

```{r handling-margins-of-error-with-6, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r handling-margins-of-error-with-6-hint-1, eval = FALSE}
moe_prop(25, ..., 5, ...)
```

```{r handling-margins-of-error-with-6-test, include = FALSE}
moe_prop(25, 100, 5, 3)
```

### 

we had an ACS estimate of 25 with a margin of error of 5 around that estimate. This allowed us the determine the margin of error around the derived proportion of 0.25. The US Census Bureau publishes formulas to calculate margin of errors with the functions [`moe_prop()`](https://rdocumentation.org/packages/tidycensus/versions/1.4.4/topics/moe_prop), [`moe_sum()`](https://www.rdocumentation.org/packages/tidycensus/versions/1.5/topics/moe_sum), and [`moe_product()`](https://www.rdocumentation.org/packages/tidycensus/versions/1.5/topics/moe_product). Click on the functions to learn more about them.   

### Exercise 7

```{r handling-margins-of-error-with-7, exercise = TRUE}

```

<button onclick = "transfer_code(this)">Copy previous code</button>

```{r handling-margins-of-error-with-7-hint-1, eval = FALSE}

```

```{r handling-margins-of-error-with-7-test, include = FALSE}

```

### 

## Summary
### 

This tutorial covered [Chapter 3: Wrangling Census data with tidyverse tools](https://walker-data.com/census-r/wrangling-census-data-with-tidyverse-tools.html) from [*Analyzing US Census Data: Methods, Maps, and Models in R*](https://walker-data.com/census-r/index.html) by Kyle Walker. You learned about the [**tidycensus**](https://walker-data.com/tidycensus/index.html) package and some of its key functions, including [`get_decennial()`](https://walker-data.com/tidycensus/articles/basic-usage.html) and [`get_acs()`](https://walker-data.com/tidycensus/articles/basic-usage.html). We discussed margins of error (MOEs) in the American Community Survey and how to wrangle and interpret MOEs appropriately.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
